---
title: "An Analysis on NCAA Woman's Volleyball Statistics"
subtitle: "Math 261A Project 1"
author: "Robert Yav"
thanks: "Project repository available at: [https://github.com/peteragao/MATH261A-project-template](https://github.com/peteragao/MATH261A-project-template)."
date: today
date-format: long
abstract: ""
format: pdf
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false
library(ggplot2)
```

# Introduction

Volleyball is a sport that requires elite athleticism, hand-eye coordination, open communication, and most importantly, the ability to combine all of these aspects into a functioning team. One fundamental concept of volleyball is the three touch play: Receive, set, and spike. There are, however, situations in which the ball will not need an additional touch to get a point. Then begs the question: Does one need to make assists to make more kills in volleyball? If so, will teams that make more assists score more points? This project examines the relationship between the number of kills in a game to the number of assists a team makes in a volleyball game.

# Data

The data set used was collected from the Division 1 Woman's NCAA Volleyball 2022-2023 season [@NCAA_Womens_Volleyball]. Division 1 is the highest division in college sport divisions, which means that the highest caliber of skill at the college level is displayed in these data. There are 344 rows and 14 variables, with each row representing a team at the Division 1 level from the 2022-2023 season. The 14 variables include: Team, Conference, region, aces_per_set, assists_per_set, team_attacks_per_set, blocks_per_set, digs_per_set, hitting_pctg, kills_per_set, opp_hitting_pctg, W, L, and win_loss_pctg. My analysis focuses on two of the variables, kills_per_set and assists_per_set.

assists_per_set - The average amount of sets, passes, or digs to a teammate that directly result in a kill per set

kills_per_set - Average amount of hits that directly result in a point per set



# Methods
We fit the simple linear regression model

$$Y_i = \beta_0 +\beta_1 X_i + \varepsilon_i$$

  $\beta_1$ represents the expected point change in kills per set per point change in assists
  per set.
  
  $\beta_0$ represents the intercept or expected number of kills per set if we had a team with 0      assists per set. In a sports organization with players playing at the highest caliber, players     will be making assists in a game, so this statistic holds no practical relevance.
  
  $X_i$ represents the $i^{th}$ team's assists as a predictor in the model.
  
  $Y_i$ represents the $i^{th}$ team's kills as a response in the model.
  
  We implement the analysis using R [@r_language]. Visual aid done in [@ggplot].

```{r}
#| warning: false
#| message: false
#| echo: false
vb_data=read.csv("volleyball_ncaa_div1_2022_23.csv")
ggplot(data=vb_data,aes(x=assists_per_set,y=kills_per_set)) +
  geom_point() +
  labs(title = "Volleyball Kills as Explained by Assists",
       subtitle= "Data from Female NCCA Teams",
       x = "Assists per Set",
       y = "Kills per Set")
```

A quick observation indicates two potential outliers and a noticeable positive
pattern between the variables. The outliers, however, under closer inspection, cannot
be reasonably removed, as the teams were just low performers compared to the other
teams in the season, which still is important to include. (i.e. 28 losses to 0 wins)

# Results
```{r}
#| warning: false
#| message: false
#| echo: false
vb_model=lm(data=vb_data,vb_data$kills_per_set~vb_data$assists_per_set)
ggplot(data=vb_data,aes(x=assists_per_set,y=kills_per_set)) +
  geom_point() +
  labs(title = "Volleyball Kills as Explained by Assists",
       subtitle= "Data from Female NCCA Teams",
       x = "Assists per Set",
       y = "Kills per Set")+
  geom_abline(slope = vb_model$coefficients[2], 
              intercept = vb_model$coefficients[1],
              color="red")
```
The slope parameter $b_1 = 1.023$, meaning that the model estimates that for each
point change in assists, the expected point total should change by 1.023.
The intercept parameter $b_0 = 0.596$, meaning that for 0 assists, the expected amount
of points a team would have would be 0.596.
$R^2 = .9819$, meaning that our model explains about 98.19% of variation in our sample.
This is a good indicator that our model is performing well.

```{r}
#| warning: false
#| message: false
#| echo: false
ggplot(data=vb_model,aes(x=vb_model$residuals,y=vb_model$fitted.values)) +
  geom_point() +
  labs(title = "Residuals vs Fitted Values",
       y = "Residuals",
       x = "Fitted Values")

```
An observation of the residuals versus the fitted values revealed that our assumptions of normality and constant variance have been broken. As we can see, most of our residuals are positive, meaning that our model generally underestimates the true number of points scored, as the observed values are generally larger than the predicted values ($Y_i >\hat{Y_i}$).

```{r}
#| warning: false
#| message: false
#| echo: false
qqnorm(vb_model$residuals)
qqline(vb_model$residuals)
```

A quantile-quantile plot of the residuals revealed a large tail at the right end,
signifying that there were more values on the extreme than expected under a normal
distribution.


# Discussion

## Limitations

We are aware that this data only covers D1 woman's NCAA athletes, which excludes all
men, recreational, a co-ed players, as well as any lower division woman's teams. Also,
we can only make statistical inferences on D1 woman's teams, since the data does not
magnify into individual players.

Our assumptions of normality and constant variance have been broken by our residual
to fitted values plot. This is potentially due to multiple factors, such as the idea that our
values are mostly on the extreme right end of the data, denoting that most of these D1-level teams seek out high-scoring athletes and have maintained D1 status due to this requirement.



## Takeaways

There is a positive correlation between the number of assists a team makes and the number of kills a team gets. In our model, each additional assist is expected to land us an additional $b_1=1.023$ points.

Although this model does have a good fit due to its high $R^2=.9819$, due to violations of normality and constant variance, we may require additional data transforming, or fitting
a different model besides simple linear regression.



# References
